---
title: "Working with dplyr"
author: "Anthony Kenny"
date: "3 May 2016"
output: 
  html_document:
    toc: true # table of content true
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

dplyr is a package that is used to make working with data and datasets easier. It consists of five main verbs:

* select: chooses subset of columns from dataset
* filter: chooses subset of rows from the dataset
* arrange: re-orders the rows in a dataset
* mutate: build new columns from data
* summarise: calculate summary statistics

# The select verb
The select verb is of the form

`select(df, var1, var2)`

You can also use : to select a range of variables and - to exclude some variables, similar to indexing a data.frame with square brackets. You can use both variable's names as well as integer indexes. This call selects the four first variables except for the second one of a data frame df:

`select(df, 1:4, -2)`

```{r eval=FALSE}
# Print out a tbl with the four columns of hflights related to delay
select(hflights, ActualElapsedTime, AirTime, ArrDelay,DepDelay)

# Print out the columns Origin up to Cancelled of hflights
select(hflights, Origin:Cancelled)

# Answer to last question: be concise!
select(hflights,1:4, 12:21)
```

## Using helper functions

dplyr comes with a set of helper functions that can help you select groups of variables inside a `select()` call:

* starts_with("X"): every name that starts with "X",
* ends_with("X"): every name that ends with "X",
* contains("X"): every name that contains "X",
* matches("X"): every name that matches "X", where "X" can be a regular expression,
* num_range("x", 1:5): the variables named x01, x02, x03, x04 and x05,
* one_of(x): every name that appears in x, which should be a character vector.


```{r eval=FALSE}
# Print out a tbl containing just ArrDelay and DepDelay
select(hflights, ArrDelay, DepDelay)

# Print out a tbl as described in the second instruction, using both helper functions and variable names
select(hflights, 7:9, starts_with("Cancell"))

# Print out a tbl as described in the third instruction, using only helper functions.
select(hflights, ends_with("Time"), ends_with("Delay"))
```

## Comparison to basic R

```{r eval=FALSE}
# Finish select call so that ex1d matches ex1r
ex1r <- hflights[c("TaxiIn", "TaxiOut", "Distance")]
ex1d <- select(hflights, 16:18)

# Finish select call so that ex2d matches ex2r
ex2r <- hflights[c("Year", "Month", "DayOfWeek", "DepTime", "ArrTime")]
ex2d <- select(hflights, 1:6, -3)

# Finish select call so that ex3d matches ex3r
ex3r <- hflights[c("TailNum", "TaxiIn", "TaxiOut")]
ex3d <- select(hflights, starts_with("T"))
```

# The mutate verb
mutate() is the second of five data manipulation functions you will get familiar with in this course. mutate() creates new columns which are added to a copy of the dataset.

Take this example that adds a new column, z, which is the element-wise sum of the columns x and y, to the data frame df:
`mutate(df, z = x + y)`


```{r eval=FALSE}
# hflights and dplyr are loaded and ready to serve you.

# Add the new variable ActualGroundTime to a copy of hflights and save the result as g1.
g1 <- mutate(hflights, ActualGroundTime = ActualElapsedTime - AirTime)

# Add the new variable GroundTime to g1. Save the result as g2.
g2 <- mutate(g1, GroundTime = TaxiIn + TaxiOut)

# Add the new variable AverageSpeed to g2. Save the result as g3.
g3 <- mutate(g2, AverageSpeed = Distance / AirTime * 60)

# Print out g3
g3
```

## Add multiple variables using Mutate
mutate() even allows you to use a new variable while creating a next variable in the same call. In this example, the new variable x is directly reused to create the new variable y:
`mutate(my_df, x = a + b, y = x + c)`

```{r eval=FALSE}
# Add a second variable loss_ratio to the dataset: m1
m1 <- mutate(hflights, loss = ArrDelay - DepDelay, loss_ratio = loss/DepDelay)

# Add the three variables as described in the third instruction: m2
m2 <- mutate(hflights, TotalTaxi = TaxiIn + TaxiOut, ActualGroundTime = ActualElapsedTime - AirTime, Diff = TotalTaxi - ActualGroundTime)
```

# The filter verb
R comes with a set of logical operators that you can use inside filter():

* x < y, TRUE if x is less than y
* x <= y, TRUE if x is less than or equal to y
* x == y, TRUE if x equals y
* x != y, TRUE if x does not equal y
* x >= y, TRUE if x is greater than or equal to y
* x > y, TRUE if x is greater than y
* x %in% c(a, b, c), TRUE if x is in the vector c(a, b, c)

The following example filters df such that only the observations for which a is positive, are kept:

`filter(df, a > 0)`

```{r eval=FALSE}
# hflights is at your disposal as a tbl, with clean carrier names

# All flights that traveled 3000 miles or more
filter(hflights, Distance > 3000)

# All flights flown by one of JetBlue, Southwest, or Delta
filter(hflights, UniqueCarrier %in% c("JetBlue","Southwest", "Delta"))

# All flights where taxiing took longer than flying
filter(hflights, TaxiIn + TaxiOut > AirTime)
```

## Combining tests using boolean operators
R also comes with a set of boolean operators that you can use to combine multiple logical tests into a single test. These include & (and), | (or), and ! (not). Instead of using the & operator, you can also pass several logical tests to filter(), separated by commas. The following two calls are completely equivalent:

`filter(df, a > 0 & b > 0)`
`filter(df, a > 0, b > 0)`
Next, is.na() will also come in handy. This example keeps the observations in df for which the variable x is not NA:

`filter(df, !is.na(x))`

```{r eval=FALSE}
# All flights that departed before 5am or arrived after 10pm
filter(hflights, DepTime < 500 | ArrTime > 2200)

# All flights that departed late but arrived ahead of schedule
filter(hflights, DepDelay > 0, ArrDelay < 0)

# All flights that were cancelled after being delayed
filter(hflights, Cancelled == 1, DepDelay > 0)
```

## Blend together what you've learned
Let's build some more complex examples

```{r eval=FALSE}
# Select the flights that had JFK as their destination: c1
c1 <- filter(hflights, Dest == "JFK")

# Combine the Year, Month and DayofMonth variables to create a Date column: c2
c2 <- mutate(c1, Date = paste(Year, Month, DayofMonth, sep = "-"))

# Print out a selection of columns of c2
select(c2, Date, DepTime, ArrTime, TailNum)
```

# The arrange verb
The arrange verb re-orders rows in a dataset

```{r eval=FALSE}
# dplyr and the hflights tbl are available

# Definition of dtc
dtc <- filter(hflights, Cancelled == 1, !is.na(DepDelay))

# Arrange dtc by departure delays
arrange(dtc, DepDelay)

# Arrange dtc so that cancellation reasons are grouped
arrange(dtc, CancellationCode)

# Arrange dtc according to carrier and departure delays
arrange(dtc, UniqueCarrier, DepDelay)
```

## Reverse the order of arrange

```{r eval=FALSE}
# Arrange according to carrier and decreasing departure delays
arrange(hflights, UniqueCarrier, desc(DepDelay))

# Arrange flights by total delay (normal order).
arrange(hflights, DepDelay + ArrDelay)
```

# The summarise verb
We can use many aggregate functions with R, some of these include:
sum, mean, median, var, IQR, sd, min, max, length
and dplyr supplies some of its own like:
first, last, nth, n,  n_distinct

```{r eval=FALSE}
# hflights and dplyr are loaded in the workspace

# Print out a summary with variables min_dist and max_dist
summarise(hflights, min_dist = min(Distance), max_dist = max(Distance))

# Print out a summary with variable max_div the max distance of a diverted flight
summarise((filter(hflights, Diverted > 0)), max_div = max(Distance))
```

## Aggregate functions

R contains many aggregating functions, as dplyr calls them:

* min(x) - minimum value of vector x.
* max(x) - maximum value of vector x.
* mean(x) - mean value of vector x.
* median(x) - median value of vector x.
* quantile(x, p) - pth quantile of vector x.
* sd(x) - standard deviation of vector x.
* var(x) - variance of vector x.
* IQR(x) - Inter Quartile Range (IQR) of vector x.
* diff(range(x)) - total range of vector x.

```{r eval=FALSE}
# hflights is available

# Remove rows that have NA ArrDelay: temp1
temp1 <- filter(hflights, !is.na(ArrDelay))

# Generate summary about ArrDelay column of temp1
summarise(temp1, earliest = min(ArrDelay), average = mean(ArrDelay), latest = max(ArrDelay), sd = sd(ArrDelay))

# Keep rows that have no NA TaxiIn and no NA TaxiOut: temp2
temp2 <- filter(hflights, !is.na(TaxiIn) & !is.na(TaxiOut))

# Print the maximum taxiing difference of temp2 with summarise()
summarise(temp2, max_taxi_diff = max(abs(TaxiIn - TaxiOut)))
```

## dplyr aggregate functions
dplyr provides several helpful aggregate functions of its own, in addition to the ones that are already defined in R. These include:

* first(x) - The first element of vector x.
* last(x) - The last element of vector x.
* nth(x, n) - The nth element of vector x.
* n() - The number of rows in the data.frame or group of observations that summarise() describes.
* n_distinct(x) - The number of unique values in vector x.

Next to these dplyr-specific functions, you can also turn a logical test into an aggregating function with sum() or mean(). A logical test returns a vector of TRUE's and FALSE's. When you apply sum() or mean() to such a vector, R coerces each TRUE to a 1 and each FALSE to a 0. sum() then represents the total number of observations that passed the test; mean() represents the proportion.

```{r eval=FALSE}
# hflights is available with full names for the carriers

# Generate summarizing statistics for hflights
summarise(hflights, n_obs = n(), n_carrier = n_distinct(UniqueCarrier), n_dest = sum(n_distinct(Dest)))

# All American Airline flights
aa <- filter(hflights, UniqueCarrier == "American")

# Generate summarizing statistics for aa , number of flights, num cancelled aa flights,
# mean arrival delay.
summarise(aa, n_flights = n(), n_canc = sum(Cancelled), avg_delay = mean(ArrDelay, na.rm=TRUE))
```


# The pipe operator
The pipe operator allows us to chain functions together.
It removes the first argument of the dplyr functions and places it on the left of the pipe `%>%`

Let's turn this english sentence into R code using the pipe operator:
* Take the hflights data set and then …
* Add a variable named diff that is the result of subtracting TaxiIn from TaxiOut, and then …
* Pick all of the rows whose diff value does not equal NA, and then …
* Summarise the data set with a value named avg that is the mean diff value.

```{r eval=FALSE}
hflights %>% mutate( diff = TaxiOut - TaxiIn) %>% filter(!is.na(diff)) %>% summarise(avg = mean(diff))
```


## Drive or fly (part 1)
Over the next few exercises you will examine whether it sometimes makes sense to drive instead of fly. You will begin by making a data set that contains relevant variables. Then, you will find flights whose equivalent average velocity is lower than the velocity when traveling by car.

* mutate() the hflights dataset and add two variables:
RealTime: the actual elapsed time plus 100 minutes (for the overhead that flying involves) and
mph: calculated as Distance / RealTime * 60, then

* filter() to keep observations that have an mph that is not NA and below 70, finally

* summarise() the result by creating four summary variables:
n_less, the number of observations,
n_dest, the number of destinations,
min_dist, the minimum distance and
max_dist, the maximum distance.

```{r eval=FALSE}
# Chain together mutate(), filter() and summarise()
hflights %>% mutate(RealTime = ActualElapsedTime + 100, mph = Distance/RealTime * 60) %>% filter( !is.na(mph) & mph < 70) %>% summarise( n_less = n(), n_dest = sum(n_distinct(Dest)), min_dist = min(Distance), max_dist = max(Distance))
```


## Drive or fly (part 2)
The previous exercise suggested that some flights might be less efficient than driving in terms of speed. But is speed all that matters? Flying imposes burdens on a traveler that driving does not. For example, airplane tickets are very expensive. Air travelers also need to limit what they bring on their trip and arrange for a pick up or a drop off. Given these burdens we might demand that a flight provide a large speed advantage over driving.

Let's define preferable flights as flights that are 150% faster than driving, i.e. that travel 105 mph or greater in real time. Also, assume that cancelled or diverted flights are less preferable than driving.
```{r eval=FALSE}
# Finish the command with a filter() and summarise() call
hflights %>%
  mutate(RealTime = ActualElapsedTime + 100, mph = Distance / RealTime * 60) %>%
  filter( mph < 105 | Cancelled == 1 | Diverted == 1) %>%
  summarise(n_non = n(), n_dest = n_distinct(Dest), min_dist = min(Distance), max_dist = max(Distance))
```

The results show that almost 19% of flights appear less desirable than simply driving to the destination, which is rather surprising!

Let's answer another question, 
How many flights where overnight flights?

```{r eval=FALSE}
hflights %>% filter( !is.na(DepTime) & !is.na(ArrTime) & DepTime > ArrTime) %>%
    summarise(num = n())
```

# Using group_by()
This function groups observations based on a particular variable, it is particularly powerful when using it with other dplyr verbs such as summarise

```{r eval=FALSE}

# Make an ordered per-carrier summary of hflights containing percent of cancelled flights &
# average delay per carrier
hflights %>% 
    group_by(UniqueCarrier) %>%
    summarise(p_canc = mean(Cancelled == 1) * 100 , avg_delay = mean(ArrDelay, na.rm = TRUE))%>%
    arrange(avg_delay)
```

## Using group_by with mutate()
You can also combine group_by() with mutate(). When you mutate grouped data, mutate() will calculate the new variables independently for each group. This is particularly useful when mutate() uses the rank() function, that calculates within-group rankings. rank() takes a group of values and calculates the rank of each value within the group. As with `arrange()`, `rank()` ranks values from the largest to the smallest.

```{r eval=FALSE}
# Ordered overview of average arrival delays per carrier
hflights %>%
    filter(ArrDelay > 0 & !is.na(ArrDelay)) %>%
    group_by(UniqueCarrier) %>%
    summarise( avg = mean(ArrDelay)) %>%
    mutate( rank = rank(avg)) %>%
    arrange(rank(rank))
```

# Exercise
Let's answer some questions about the flights dataset like:
* How many airplanes only flew to one destination from Houston?
* Find the most visited destination for each carrier?

```{r eval=FALSE}
# How many airplanes only flew to one destination from Houston?
hflights %>%
  group_by(TailNum) %>%
  summarise(ndest = n_distinct(Dest)) %>%
  filter(ndest == 1) %>%
  summarise(nplanes = n())

# Find the most visited destination for each carrier?
hflights %>% 
  group_by(UniqueCarrier, Dest) %>%
  summarise(n = n()) %>%
  mutate(rank = rank(desc(n))) %>%
  filter(rank == 1)
```

## dplyr with data.table
dplyr can also deal with data.table datasets.
Example:

```{r eval=FALSE}
#Convert hflights to a data.table
library(data.table)
hflights2 <- as.data.table(hflights)

# Use summarise to calculate n_carrier
hflights2 %>% summarise( n_carrier = n_distinct(UniqueCarrier))
```

# dplyr and mySQL databases
DataCamp hosts a mySQL database with data about flights that departed from New York City in 2013. The data is similar to the data in hflights, but it does not contain information about cancellations or diversions. With the tbl() function, we already created a reference to a table in this information.

Although nycflights is a reference to data that lives outside of R, you can use the dplyr commands on them as usual. Behind the scenes, dplyr will convert the commands to the database's native language (in this case, SQL), and return the results. This allows you to pull data that is too large to fit in R: only the fraction of the data that you need will actually be downloaded into R, which will usually fit into R without memory issues.

To learn all about dplyr and databases, visit the dplyr database vignette.

Try to understand the code that creates nycflights, a reference to a MySQL table.
Use glimpse() to check out nycflights. Although nycflights is a reference to a tbl in a remote database, there is no difference in syntax. Look carefully: the variable names in nycflights differ from the ones in hflights!
Group nycflights data by carrier, then summarise() with two variables: n_flights, the number of flights flown by each carrier and avg_delay, the average arrival delay of flights flown by each carrier. Finally, arrange the carriers by average delay from low to high.

```{r eval=FALSE}
# Set up a connection to the mysql database
my_db <- src_mysql(dbname = "dplyr", 
                   host = "dplyr.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                   port = 3306, 
                   user = "dplyr",
                   password = "dplyr")

# Reference a table within that source: nycflights
nycflights <- tbl(my_db, "dplyr")

# glimpse at nycflights
glimpse(nycflights)

# Ordered, grouped summary of nycflights
  nycflights %>%
    group_by(carrier) %>%
    summarise(n_flights = n(), avg_delay = mean(arr_delay)) %>%
    arrange(avg_delay)
```

Great, we're done for now!
